### 2020-07-29 manuscript.tex
tidy ups

The molecular clock hypothesis states that the evolutionary rates of biological sequences are approximately constant throughout time \cite{zuckerkandl1962molecular}.
This assumption forms the basis of phylogenetics, under which evolutionary trees and divergence dates of life forms are inferred from biological sequences, such as nucleotides and amino acids \cite{douzery2003local, drummond2006relaxed}.
In Bayesian phylogenetics, these trees and their associated parameters are estimated as probabilitiy distributions \cite{kuhner1995estimating, larget1999markov, mau1999bayesian}. 
Statistical inference is typically performed by sampling using the Markov chain Monte Carlo (MCMC) algorithm \cite{metropolis53, hastings70}.
Bayesian phylogenetic clock models have been implemented in many platforms including BEAST \cite{drummond2012bayesian}, BEAST2 \cite{bouckaert2019beast}, MrBayes \cite{ronquist2012mrbayes}, and RevBayes \cite{hohna2016revbayes}.

The simplest phylogenetic clock model -- the strick clock -- makes the convenient assumption that the evolutionary rate is constant across all lineages \cite{zuckerkandl1965evolutionary, kuhner1995estimating, larget1999markov}.  % Early examples? 
However, molecular substitution rates are known to vary over time, over population sizes, over evolutionary pressures, and over DNA replicative machineries \cite{gillespie1994causes, woolfit2009effective, loh2010optimization}.
Moreover, any given dataset may be clock-like (where the substitution rate has a small variance across lineages) or non clock-like (a large variance). 
In the latter case, a strict clock is probably not suitable.

This led to the development of relaxed (uncorrelated) clock models, under which each branch in the phylogenetic tree has its own substitution rate  \cite{drummond2006relaxed}.
Branch rates can be drawn from a range of probability distributions including log-normal, exponential, gamma, and inverse-gamma distributions \cite{drummond2006relaxed, lepage2007general, li2012model}.
This class of models is widely used, and has aided insight into many recent biological problems, including the 2016 Zika virus outbreak \cite{faria2017establishment} and the COVID-19 pandemic \cite{giovanetti2020first}.
% is compatible with increasingly prominent multispecies coalescent models \cite{ogilvie2017starbeast2},

@@

Let $t_i$ be the height (time) of node $i$.
Each node $i$ in $\mathcal{T}$, except for the root, is associated with a parental branch length $\tau_i$ (the height difference between $i$ and its parent)  and a parental branch substitution rate $r_i = r(\mathcal{R}_i)$. 
In a relaxed clock model, each of the $2N-2$ elements in $\mathcal{R}_i$ are independently distributed under the prior $p(\vec{\mathcal{R}}^{\,} | \sigma)$.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/rateparameterisation.pdf}
\caption{\textbf{Branch rate parameterisations.} Top left: the branch rate prior under a log-normal($-0.5\sigma^2, \sigma$) distribution (with a mean of 1). 
The function for transforming $\mathcal{R}$ into branch rates $r(\mathcal{R})$ is depicted for \textit{real} (top right), \textit{cat} (bottom left), and \textit{quant} (bottom right).
For simplicity there are 10 \textit{cat} bins displayed, however in practice there are $2N-2$ bins. }
\label{fig:rateparams}
\end{figure}

@@

Under the prior distribution $p(\vec{\mathcal{R}}^{\,} | \sigma)$, rates are log-normally distributed with a mean of 1:

@@

where $\mu = -0.5\sigma^2$ is set such that the expected value of the log-normal distribution is 1.
In this article we only consider log-normal clock priors, however the methods discussed are general.

Zhang and Drummond 2020 present a series of tree operators which propose internal/root node heights, and then recompute the rates of incident branches such that their genetic distances ($r_i \times \tau_i$) remain constant after the proposal \cite{zhang2020improving}. 
These operators account for the correlation which exists between branch rates and branch times.
By keeping the genetic distance of each branch constant, the likelihood is unaltered by the proposal.

@@

\subsubsection*{2. Categories}
The category parameterisation (\textit{cat}) is an abstraction of the \textit{real} parameterisation. Each branch is assigned an integer from $0$ to $n-1$:

@@

These integers correspond to $n$ rate categories.
 The domain of $\vec{\mathcal{R}}^{\,}$ is uniformly distributed:

@@

Let $f(x|\sigma)$ be the probability density function (pdf) and let $F(x|\sigma) = \int\limits_{0}^{x} f(t|\sigma) \; dt$ be the cumulative distribution function (cdf) of the prior distribution used by the underlying \textit{real} clock model. Then, in the \textit{cat} parameterisation, $f(x|\sigma)$ is discretised into $n$ bins and the elements of $\vec{\mathcal{R}}^{\,}$ each point to one of these bins. Each bin contains uniform probability density $1/n$. The rate of each bin is equal to the median value within the bin

@@

where $F^{-1}$ is the inverse cumulative distribution function (i-cdf).

The key advantage of the \textit{cat} parameterisation is the removal of a term from the posterior density (Equation \ref{eq:bayesian}), or more accurately the replacement of a non-trivial $p(\vec{\mathcal{R}}^{\,} | \sigma)$ term with that of a uniform prior. Thus, one fewer term needs to be estimated per rate.

@@

Finally, rates can be parameterised as real numbers $0 < \mathcal{R}_i < 1$ which describe the rate's quantile with respect to some underlying clock model distribution. Under the  \textit{quant} parameterisation, each element in $\vec{\mathcal{R}}^{\,}$ is uniformly distributed.

@@

Transforming these quantiles into rates invokes the i-cdf of the underlying \textit{real} clock model distribution:

@@

While this approach has clear similarities with \textit{cat}, the domain of rates here is continuous (as opposed to being confined to a discrete number of bins).
In this project we extended the family of constant distance operators  \cite{zhang2020improving} so they are compatible with \textit{quant}.
These are presented in \textbf{\nameref{sect:S1Appendix}}.

@@

The weight of an operator determines the probability of the operator being selected at each step during MCMC.
In BEAST2, operators can have a tunable parameter $s$ which determines the step size of the operator \cite{bouckaert2019beast}. 
Although $s$ is learned throughout the MCMC chain, the operator weights themselves are typically held constant.
Pre-existing BEAST2 clock model operators (i.e. those which generate proposals for either $\mathcal{R}$ or $\sigma$) are summarised in \textbf{Table \ref{table:kernels}}, and further operators are introduced throughout the paper.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l| l p{4cm} l l|} 
 \hline
 & Operator & Description & Parameters & Parameterisations  \\
  \hline
 1 & \texttt{RandomWalk} & Moves a single element by a tunable amount. & $\mathcal{R}, \sigma$ & \textit{cat}, \textit{real}, \textit{quant} \\
  \hline
 2 & \texttt{Scale} & Applies \texttt{RandomWalk} on the log-transformation (suitable for parameters with positive domains). & $\mathcal{R}, \sigma$ & \textit{real}, \textit{quant}  \\
  \hline
 3 & \texttt{Interval} & Applies \texttt{RandomWalk} on the logit-transformation (suitable for parameters with upper and lower limits). & $\mathcal{R}$ & \textit{quant}  \\
  \hline
   4 & \texttt{Swap} & Swaps two random elements in the vector \cite{drummond2006relaxed}. & $\mathcal{R}$  & \textit{cat}, \textit{real}, \textit{quant}  \\
 \hline
  5 & \texttt{Uniform} & Resamples one element in the vector from a uniform distribution. & $\mathcal{R}$  & \textit{cat}, \textit{quant}  \\
 \hline
 6 & \texttt{ConstantDistance} & Adjusts an internal node and recalculates all incident branch rates such that the genetic distances remain constant \cite{zhang2020improving}.  & $\mathcal{R}, \mathcal{T}$ & \textit{real}, \textit{quant} \\
 \hline
 7 & \texttt{SimpleDistance} & Applies \texttt{SimpleDistance} to the root node \cite{zhang2020improving}.  & $\mathcal{R}, \mathcal{T}$ & \textit{real}, \textit{quant} \\
 \hline
 8 & \texttt{SmallPulley} & Proposes new branch rates incident to the root such that their combined genetic distance is constant  \cite{zhang2020improving}.  & $\mathcal{R}$ & \textit{real}, \textit{quant} \\
 \hline
 9 & \texttt{FastClockScale} & Applies \texttt{Scale} to $\sigma$ and then recomputes all rates $\mathcal{R}$ such that their quantiles under the prior $p(\mathcal{R}|\sigma)$ are constant \cite{zhang2020improving}.  & $\mathcal{R}$ & \textit{real} \\
 \hline
\end{tabular}
\caption{Summary of pre-existing BEAST2 operators which apply to either branch rates $\mathcal{R}$ or the clock standard deviation $\sigma$, and the substitution rate parameterisation they apply to.
 \texttt{ConstantDistance} and \texttt{SimpleDistance} also adjust tree node heights. }
\label{table:kernels}
\end{table}

@@

We hypothesise that datasets with strong signal (or large $L$) will mix best when the more precise and meticulous kind of operator is employed, such as those informed by correlations within the posterior distribution e.g. the family of constant distance operators \cite{zhang2020improving}.
Whereas, datasets which contain very poor signal (or small $L$) are likely to mix best when there is more weight placed on bold operators such as \texttt{SampleFromPrior}, \texttt{Swap}, or \texttt{Uniform} which sample each branch rate independently of its current estimate.
\texttt{AdaptiveOperatorSampler} should learn the combination of weights behind these classes of operators best suited for the dataset.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l p{4cm}|} 
 \hline
 Meta-operator & Operators \\
\hline
 \multirow{4}{*}{\texttt{AdaptiveOperatorSampler($\sigma$)}} & \texttt{FastClockScale($\sigma, \mathcal{R}$)}$^{*}$ \\ 
 & \texttt{RandomWalk($\sigma$)} \\
 & \texttt{Scale($\sigma$)} \\
 & \texttt{SampleFromPrior($\sigma$)} \\
 \hline
  \multirow{6}{*}{\texttt{AdaptiveOperatorSampler($\mathcal{R}$)}} & \texttt{ConstantDistance($\mathcal{R}, \mathcal{T}$)} \\ 
& \texttt{RandomWalk($\mathcal{R}$)} \\
& \texttt{Scale($\mathcal{R}$)}$^{*}$ \\
& \texttt{Interval($\mathcal{R}$)}$^{**}$  \\
& \texttt{Swap($\mathcal{R}$)} \\
& \texttt{SampleFromPrior($\mathcal{R}$)} \\
 \hline
   \multirow{2}{*}{\texttt{AdaptiveOperatorSampler(root)}} & \texttt{SimpleDistance($\mathcal{R}, \mathcal{T}$)} \\ 
&  \texttt{SmallPulley($\mathcal{R}, t$)}\\
 \hline
\end{tabular}
\caption{Summary of the three instances of \texttt{AdaptiveOperatorSampler} used under the \textit{real} and \textit{quant} parameterisations.  $^{*}$\textit{real} only. $^{**}$\textit{quant} only. }
\label{table:adaptiveSampling}
\end{table}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/bactrian.pdf}
\caption{\textbf{The Bactrian proposal kernel.} Y-axis corresponds to probability density $f(\Sigma|m)$.  }
\label{fig:bactrian}
\end{figure}

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l| p{3cm} l l|} 
 \hline
 & Operator(s) & Proposal & Parameter $x$   \\
  \hline
 1 & \texttt{RandomWalk}  & $x^\prime \leftarrow x + s\Sigma$ & $\mathcal{R}, \sigma$  \\
  \hline
 2 & \texttt{Scale} & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\mathcal{R}, \sigma$   \\
  \hline
 3 & \texttt{Interval} & $\begin{array} {rl} &y \leftarrow \frac{1 - x}{x} \times e^{s\Sigma} \\ &x^\prime \leftarrow \frac{y}{y + 1}  \end{array}$ & $\mathcal{R}$  \\
  \hline
 4 & \texttt{ConstantDistance} \texttt{SimpleDistance} & $x^\prime \leftarrow x + s\Sigma$ & $t$ \\
 \hline
 5 & \texttt{SmallPulley} & $x^\prime \leftarrow x + s\Sigma$ & $\mathcal{R}$  \\
 \hline
6 & \texttt{FastClockScale}  & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\sigma$   \\
 \hline
\end{tabular}
\caption{Proposal kernels $q(x^\prime|x)$ of clock model operators.
 In each operator, $\Sigma$ is drawn from either a $\text{Bactrian}(m)$ or $\text{Uniform}$ distribution (distributions are normalised so that they have a mean of 0 and a variance of 1). 
 The scale size $s$ is tunable.
 \texttt{ConstantDistance} and \texttt{SimpleDistance} propose tree heights $t$.
  The \texttt{Interval} operator applies to rate quantiles and respects its domain i.e. $0 < x, x^\prime < 1$. }
\label{table:bactriankernels}
\end{table}

@@

The AVMVN kernel assumes its parameters live in $x \in \mathbb{R}^N$ and that these parameters follow a multivariate normal distribution with covariance matrix $\Sigma_N$. Hence, the kernel operates on the logarithmic or logistic transformation of the $N$ leaf branch rates, depending on the rate parameterisation:

@@

where $\mathcal{MVN}$ is the multivariate normal probability density. $\beta \; (= 0.05)$ is a constant which determines the fraction of the proposal determined by the identity matrix $\mathbb{I}_N$, as opposed to the covariance matrix $\Sigma_D$ which is trained during MCMC.

@@

Throughout \nameref{sect:models}, we introduced four new operators which are applied to the relaxed clock. 
These operators are summarised in \textbf{Table \ref{table:newOperators}} and are applied in conjunction with the pre-existing clock model operators presented in \textbf{Table \ref{table:kernels}}.

\begin{table}[h!]
\centering
\begin{tabular}{|l| l p{4cm} l|} 
 \hline
 & Operator & Description & Parameters  \\
  \hline
1 & \texttt{AdaptiveOperatorSampler} & Samples sub-operators proportionally to their weights, which are learned (see \nameref{sect:adaptiveSampling}). & $\mathcal{R}, \sigma, \mathcal{T}$ \\
  \hline
2 & \texttt{SampleFromPrior} & Resamples a random number of elements from their prior (see \nameref{sect:adaptiveSampling}). & $\mathcal{R}, \sigma$ \\
  \hline
3 & \texttt{NarrowExchangeRate} & Moves a branch and recomputes branch rates so that genetic distances are constant (see \nameref{sect:NER}). & $\mathcal{R}, \mathcal{T}$\\
  \hline
4 & \texttt{LeafAVMVN}  & Proposals new rates for all leaves in one move (see \nameref{AVMVN_sect}) \cite{baele2017adaptive}. & $\mathcal{R}$ \\
  \hline
\end{tabular}
\caption{Summary of clock model operators introduced throughout this article.}
\label{table:newOperators}
\end{table}

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l l l|} 
 \hline
Configuration & Operator & Weight  \\
  \hline
 \multirow{4}{*}{\textit{cat}} 
 	& \texttt{RandomWalk($\mathcal{R}$)} & $10$ \\
 	& \texttt{Uniform($\mathcal{R}$)} & $10$  \\ 
 	& \texttt{Swap($\mathcal{R}$)} & $10$ \\ 
 	& \texttt{Scale($\sigma$)} & $10$ \\ 
  \hline
\end{tabular}
\caption{Operator configurations for \textit{cat} in Round 2. The configurations of \textit{real} (adapt) and \textit{quant} (adapt) are shown in \textbf{Table \ref{table:round1Results}}.}
\label{table:round2Results}
\end{table}

>>>>

The molecular clock hypothesis states that the evolutionary rates of biological sequences are approximately constant throughout time \cite{zuckerkandl1962molecular}.
This assumption forms the basis of phylogenetics, under which the evolutionary trees and divergence dates of life forms are inferred from biological sequences, such as nucleotides and amino acids \cite{douzery2003local, drummond2006relaxed}.
In Bayesian phylogenetics, these trees and their associated parameters are estimated as probabilitiy distributions \cite{kuhner1995estimating, larget1999markov, mau1999bayesian}. 
Statistical inference is typically performed by sampling using the Markov chain Monte Carlo (MCMC) algorithm \cite{metropolis53, hastings70}.
Bayesian phylogenetic clock models have been implemented in many platforms including BEAST \cite{drummond2012bayesian}, BEAST2 \cite{bouckaert2019beast}, MrBayes \cite{ronquist2012mrbayes}, and RevBayes \cite{hohna2016revbayes}.

The simplest phylogenetic clock model -- the strick clock -- makes the convenient assumption that the evolutionary rate is constant across all lineages \cite{zuckerkandl1965evolutionary, kuhner1995estimating, larget1999markov}.  % Early examples? 
However, molecular substitution rates are known to vary over time, over population sizes, over evolutionary pressures, and over nucleic acid replicative machineries \cite{gillespie1994causes, woolfit2009effective, loh2010optimization}.
Moreover, any given dataset may be clock-like (where the substitution rate has a small variance across lineages) or non clock-like (a large variance). 
In the latter case, a strict clock is probably not suitable.

This led to the development of relaxed (uncorrelated) clock models, under which each branch in the phylogenetic tree has its own molecular substitution rate  \cite{drummond2006relaxed}.
Branch rates can be drawn from a range of probability distributions including Log-Normal, Exponential, Gamma, and Inverse-Gamma distributions \cite{drummond2006relaxed, lepage2007general, li2012model}.
This class of models is widely used, and has aided insight into many recent biological problems, including the 2016 Zika virus outbreak \cite{faria2017establishment} and the COVID-19 pandemic \cite{giovanetti2020first}.
% is compatible with increasingly prominent multispecies coalescent models \cite{ogilvie2017starbeast2},

@@

Let $t_i$ be the height (time) of node $i$.
Each node $i$ in $\mathcal{T}$, except for the root, is associated with a parental branch length $\tau_i$ (the height difference between $i$ and its parent)  and a parental branch substitution rate $r_i = r(\mathcal{R}_i)$. 
In a relaxed clock model, each of the $2N-2$ elements in $\vec{\mathcal{R}}^{\,}$ are independently distributed under the prior $p(\vec{\mathcal{R}}^{\,} | \sigma)$.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/rateparameterisation.pdf}
\caption{\textbf{Branch rate parameterisations.} Top left: the prior density of a branch rate $r$ under a Log-Normal($-0.5\sigma^2, \sigma$) distribution (with its mean fixed at 1). 
The function for transforming $\mathcal{R}$ into branch rates $r(\mathcal{R})$ is depicted for \textit{real} (top right), \textit{cat} (bottom left), and \textit{quant} (bottom right).
For simplicity there are 10 \textit{cat} bins displayed, however in practice there are $2N-2$ bins. }
\label{fig:rateparams}
\end{figure}

@@

Under the prior distribution $p(\vec{\mathcal{R}}^{\,} | \sigma)$, rates are Log-Normally distributed with a mean of 1:

@@

where $\mu = -0.5\sigma^2$ is set such that the expected value of the Log-Normal distribution is 1.
In this article we only consider Log-Normal clock priors, however the methods discussed are general.

Zhang and Drummond 2020 introduced a series of tree operators which propose node heights and branch rates, such that the resulting genetic distances ($r_i \times \tau_i$) remain constant \cite{zhang2020improving}. 
These operators account for correlations between branch rates and branch times.
By keeping the genetic distance of each branch constant, the likelihood is unaltered by the proposal.

@@

\subsubsection*{2. Categories}
The category parameterisation (\textit{cat}) is an abstraction of the \textit{real} parameterisation. 
Each of the $2N-2$ branches are assigned an integer from $0$ to $n-1$:

@@

These integers correspond to $n$ rate categories (\textbf{Fig. \ref{fig:rateparams}}).
The domain of $\vec{\mathcal{R}}^{\,}$ is uniformly distributed under thr prior:

@@

Let $f(x|\sigma)$ be the probability density function (PDF) and let $F(x|\sigma) = \int\limits_{0}^{x} f(t|\sigma) \; dt$ be the cumulative distribution function (CDF) of the prior distribution used by the underlying \textit{real} clock model. 
Then, in the \textit{cat} parameterisation, $f(x|\sigma)$ is discretised into $n$ bins and each element within $\vec{\mathcal{R}}^{\,}$ points to one such bin,
where each bin has uniform prior density.
 The rate of each bin is equal to the median value within the bin

@@

where $F^{-1}$ is the inverse cumulative distribution function (i-CDF).

The key advantage of the \textit{cat} parameterisation is the removal of a term from the posterior density (Equation \ref{eq:bayesian}), or more accurately the replacement of a non-trivial $p(\vec{\mathcal{R}}^{\,} | \sigma)$ term with that of a uniform prior. 
This may facilitate efficient traversal of the parameter space by MCMC.

@@

Finally, rates can be parameterised as real numbers $0 < \mathcal{R}_i < 1$ which describe the rate's quantile with respect to some underlying clock model distribution. Under the  \textit{quant} parameterisation, each of the $2N-2$ elements in $\vec{\mathcal{R}}^{\,}$ are uniformly distributed.

@@

Transforming these quantiles into rates invokes the i-CDF of the underlying \textit{real} clock model distribution:

@@

While this approach has clear similarities with \textit{cat}, the domain of rates here is continuous (as opposed to being confined to a discrete number of bins).
In this project we extended the family of constant distance operators  \cite{zhang2020improving} so they are compatible with \textit{quant}.
These operators are presented in \textbf{\nameref{sect:S1Appendix}}.

The computational requirements of i-CDF evaluation may hinder the performance of the \textit{quant} parameterisation.

@@

The weight of an operator determines the probability of the operator being selected at each step during MCMC.
In BEAST2, operators can have a tunable parameter $s$ which determines the step size of the operator \cite{bouckaert2019beast}. 
Although $s$ is learned throughout the MCMC chain, the operator weights themselves are typically held constant.
Pre-existing BEAST2 clock model operators (i.e. those which generate proposals for either $\vec{\mathcal{R}}^{\,}$ or $\sigma$) are summarised in \textbf{Table \ref{table:kernels}}, and further operators are introduced throughout the paper.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l p{4.2cm} l l|} 
 \hline
 Operator & Description & Parameters & Parameterisations  \\
  \hline
 \texttt{RandomWalk} & Moves a single element by a tunable amount. & $\vec{\mathcal{R}}^{\,}, \sigma$ & \textit{cat}, \textit{real}, \textit{quant} \\
  \hline
\texttt{Scale} & Applies \texttt{RandomWalk} on the log-transformation (suitable for parameters with positive domains). & $\vec{\mathcal{R}}^{\,}, \sigma$ & \textit{real}, \textit{quant}  \\
  \hline
 \texttt{Interval} & Applies \texttt{RandomWalk} on the logit-transformation (suitable for parameters with upper and lower limits). & $\vec{\mathcal{R}}^{\,}$ & \textit{quant}  \\
  \hline
 \texttt{Swap} & Swaps two random elements in the vector \cite{drummond2006relaxed}. & $\vec{\mathcal{R}}^{\,}$  & \textit{cat}, \textit{real}, \textit{quant}  \\
 \hline
\texttt{Uniform} & Resamples one element in the vector from a uniform distribution. & $\vec{\mathcal{R}}^{\,}$  & \textit{cat}, \textit{quant}  \\
 \hline
\texttt{ConstantDistance} & Adjusts an internal node and recalculates all incident branch rates such that the genetic distances remain constant \cite{zhang2020improving}.  & $\vec{\mathcal{R}}^{\,}, \mathcal{T}$ & \textit{real}, \textit{quant} \\
 \hline
\texttt{SimpleDistance} & Applies \texttt{ConstantDistance} to the root node \cite{zhang2020improving}.  & $\vec{\mathcal{R}}^{\,}, \mathcal{T}$ & \textit{real}, \textit{quant} \\
 \hline
\texttt{SmallPulley} & Proposes new branch rates incident to the root such that their combined genetic distance is constant  \cite{zhang2020improving}.  & $\vec{\mathcal{R}}^{\,}$ & \textit{real}, \textit{quant} \\
 \hline
\texttt{FastClockScale} & Applies \texttt{Scale} to $\sigma$ and then recomputes all rates $\vec{\mathcal{R}}^{\,}$ such that their quantiles under the prior $p(\vec{\mathcal{R}}^{\,}|\sigma)$ are constant \cite{zhang2020improving}.  & $\mathcal{R}$ & \textit{real} \\
 \hline
\end{tabular}
\caption{Summary of pre-existing BEAST2 operators, which apply to either branch rates $\vec{\mathcal{R}}^{\,}$ or the clock standard deviation $\sigma$, and the substitution rate parameterisation they apply to.
 \texttt{ConstantDistance} and \texttt{SimpleDistance} also adjust node heights in the tree $\mathcal{T}$. }
\label{table:kernels}
\end{table}

@@

We hypothesise that datasets with strong signal (or large $L$) will mix best when the more precise and meticulous kind of operator is employed, such as those informed by correlations within the posterior distribution e.g. the family of constant distance operators \cite{zhang2020improving} (\textbf{Fig. \ref{fig:rateparams}}).
Whereas, datasets which contain very poor signal (or small $L$) are likely to mix best when there is more weight placed on bold operators such as \texttt{SampleFromPrior}, \texttt{Swap}, or \texttt{Uniform} which sample each branch rate independently of its current estimate.
\texttt{AdaptiveOperatorSampler} should learn the combination of weights behind these classes of operators best suited for any given dataset.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/correlations.pdf}
\caption{\textbf{Traversing likelihood space.}
The z-axes above are the log-likelihoods of the genetic distance $r \times \tau$ between two simulated nucleic acid sequences of length $L$, under the Jukes-Cantor substitution model \cite{jukes1969evolution}. 
Two possible proposals from the current state (red circle) are depicted.
These proposals are generated by the \texttt{RandomWalk} (\texttt{RW}) and \texttt{ConstantDistance} (\texttt{CD}) operators.
In the low signal dataset ($L=0.1$kb), both operators can traverse the likelihood space effectively.
 However, the same exact same proposal by \texttt{RandomWalk} incurs a much larger likelihood penalty in the $L=0.5$kb dataset by ``falling off the ridge'', in contrast to \texttt{ConstantDistance} which ``walks along the ridge''.
 This discrepancy is even stronger for larger datasets and thus necessitates the use of operators such as \texttt{ConstantDistance} which account for correlations between branch lengths and rates. }
\label{fig:rateparams}
\end{figure}

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l p{4cm}|} 
 \hline
 Meta-operator & Operators \\
\hline
 \multirow{4}{*}{\texttt{AdaptiveOperatorSampler($\sigma$)}} & \texttt{FastClockScale($\sigma, \vec{\mathcal{R}}^{\,}$)}$^{*}$ \\ 
 & \texttt{RandomWalk($\sigma$)} \\
 & \texttt{Scale($\sigma$)} \\
 & \texttt{SampleFromPrior($\sigma$)} \\
 \hline
  \multirow{6}{*}{\texttt{AdaptiveOperatorSampler($\vec{\mathcal{R}}^{\,}$)}} & \texttt{ConstantDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)} \\ 
& \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}$)} \\
& \texttt{Scale($\vec{\mathcal{R}}^{\,}$)}$^{*}$ \\
& \texttt{Interval($\vec{\mathcal{R}}^{\,}$)}$^{**}$  \\
& \texttt{Swap($\vec{\mathcal{R}}^{\,}$)} \\
& \texttt{SampleFromPrior($\vec{\mathcal{R}}^{\,}$)} \\
 \hline
   \multirow{2}{*}{\texttt{AdaptiveOperatorSampler(root)}} & \texttt{SimpleDistance($\vec{\mathcal{R}}^{\,}, \mathcal{T}$)} \\ 
&  \texttt{SmallPulley($\vec{\mathcal{R}}^{\,}, t$)}\\
 \hline
\end{tabular}
\caption{Summary of the three instances of \texttt{AdaptiveOperatorSampler} used under the \textit{real} and \textit{quant} parameterisations.  $^{*}$\textit{real} only. $^{**}$\textit{quant} only. }
\label{table:adaptiveSampling}
\end{table}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/bactrian.pdf}
\caption{\textbf{The Bactrian proposal kernel.} The step size made under a Bactrian proposal kernel is equal to $s \Sigma$ where $\Sigma$ is drawn from the above distribution and $s$ is tunable.   }
\label{fig:bactrian}
\end{figure}

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l| p{3cm} l l|} 
 \hline
 & Operator(s) & Proposal & Parameter $x$   \\
  \hline
 1 & \texttt{RandomWalk}  & $x^\prime \leftarrow x + s\Sigma$ & $\vec{\mathcal{R}}^{\,}, \sigma$  \\
  \hline
 2 & \texttt{Scale} & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\vec{\mathcal{R}}^{\,}, \sigma$   \\
  \hline
 3 & \texttt{Interval} & $\begin{array} {rl} &y \leftarrow \frac{1 - x}{x} \times e^{s\Sigma} \\ &x^\prime \leftarrow \frac{y}{y + 1}  \end{array}$ & $\vec{\mathcal{R}}^{\,}$  \\
  \hline
 4 & \texttt{ConstantDistance} \texttt{SimpleDistance} & $x^\prime \leftarrow x + s\Sigma$ & $t$ \\
 \hline
 5 & \texttt{SmallPulley} & $x^\prime \leftarrow x + s\Sigma$ & $\vec{\mathcal{R}}^{\,}$  \\
 \hline
6 & \texttt{FastClockScale}  & $x^\prime \leftarrow x \times e^{s\Sigma}$ & $\sigma$   \\
 \hline
\end{tabular}
\caption{Proposal kernels $q(x^\prime|x)$ of clock model operators.
 In each operator, $\Sigma$ is drawn from either a $\text{Bactrian}(m)$ or $\text{Uniform}$ distribution. 
 The scale size $s$ is tunable.
 \texttt{ConstantDistance} and \texttt{SimpleDistance} propose tree heights $t$.
  The \texttt{Interval} operator applies to rate quantiles and respects its domain i.e. $0 < x, x^\prime < 1$. }
\label{table:bactriankernels}
\end{table}

@@

The AVMVN kernel assumes its parameters live in $x \in \mathbb{R}^N$ and that these parameters follow a Multivariate Normal distribution with covariance matrix $\Sigma_N$. Hence, the kernel operates on the logarithmic or logistic transformation of the $N$ leaf branch rates, depending on the rate parameterisation:

@@

where $\mathcal{MVN}$ is the Multivariate Normal probability density. $\beta \; (= 0.05)$ is a constant which determines the fraction of the proposal determined by the identity matrix $\mathbb{I}_N$, as opposed to the covariance matrix $\Sigma_D$ which is trained during MCMC.

@@

In this article we have introduced four new clock model operators. 
These operators are summarised in \textbf{Table \ref{table:newOperators}} and are applied in conjunction with the pre-existing clock model operators presented in \textbf{Table \ref{table:kernels}}.

\begin{table}[h!]
\centering
\begin{tabular}{|l p{4cm} l|} 
 \hline
 Operator & Description & Parameters  \\
  \hline
 \texttt{AdaptiveOperatorSampler} & Samples sub-operators proportionally to their weights, which are learned (see \nameref{sect:adaptiveSampling}). & $\vec{\mathcal{R}}^{\,}, \sigma, \mathcal{T}$ \\
  \hline
 \texttt{SampleFromPrior} & Resamples a random number of elements from their prior (see \nameref{sect:adaptiveSampling}). & $\vec{\mathcal{R}}^{\,}, \sigma$ \\
  \hline
 \texttt{NarrowExchangeRate} & Moves a branch and recomputes branch rates so that genetic distances are constant (see \nameref{sect:NER}). & $\vec{\mathcal{R}}^{\,}, \mathcal{T}$\\
  \hline
 \texttt{LeafAVMVN}  & Proposals new rates for all leaves in one move (see \nameref{AVMVN_sect}) \cite{baele2017adaptive}. & $\vec{\mathcal{R}}^{\,}$ \\
  \hline
\end{tabular}
\caption{Summary of clock model operators introduced throughout this article.}
\label{table:newOperators}
\end{table}

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l l l|} 
 \hline
Configuration & Operator & Weight  \\
  \hline
 \multirow{4}{*}{\textit{cat}} 
 	& \texttt{RandomWalk($\vec{\mathcal{R}}^{\,}$)} & $10$ \\
 	& \texttt{Uniform($\vec{\mathcal{R}}^{\,}$)} & $10$  \\ 
 	& \texttt{Swap($\vec{\mathcal{R}}^{\,}$)} & $10$ \\ 
 	& \texttt{Scale($\sigma$)} & $10$ \\ 
  \hline
\end{tabular}
\caption{Operator configurations for \textit{cat} in Round 2. The configurations of \textit{real} (adapt) and \textit{quant} (adapt) are shown in \textbf{Table \ref{table:round1Results}}.}
\label{table:round2Results}
\end{table}
